---
title: "Data Science Capstone - Milestone Report"
author: "Luis Ruiz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

## Overview

This report presents an exploratory analysis of the Coursera SwiftKey dataset used for building a text prediction model. The dataset consists of three files in English: blogs, news, and Twitter. The goal is to understand the basic structure and content of the data before building the prediction algorithm and Shiny app.

---

## 1. Basic File Statistics

```{r load_data}
library(stringi)
library(ggplot2)
library(knitr)

set.seed(42)

blogs_path   <- "final/en_US/en_US.blogs.txt"
news_path    <- "final/en_US/en_US.news.txt"
twitter_path <- "final/en_US/en_US.twitter.txt"

blogs   <- readLines(blogs_path,   encoding = "UTF-8", skipNul = TRUE)
news    <- readLines(news_path,    encoding = "UTF-8", skipNul = TRUE)
twitter <- readLines(twitter_path, encoding = "UTF-8", skipNul = TRUE)
```

```{r summary_table}
file_sizes <- c(
  round(file.info(blogs_path)$size   / 1e6, 1),
  round(file.info(news_path)$size    / 1e6, 1),
  round(file.info(twitter_path)$size / 1e6, 1)
)

line_counts <- c(length(blogs), length(news), length(twitter))

word_counts <- c(
  sum(stri_count_words(blogs)),
  sum(stri_count_words(news)),
  sum(stri_count_words(twitter))
)

avg_words <- round(word_counts / line_counts, 1)

summary_df <- data.frame(
  File        = c("Blogs", "News", "Twitter"),
  `Size (MB)` = file_sizes,
  Lines       = formatC(line_counts, format = "d", big.mark = ","),
  `Total Words` = formatC(word_counts, format = "d", big.mark = ","),
  `Avg Words/Line` = avg_words,
  check.names = FALSE
)

kable(summary_df, align = "lrrrr", caption = "Table 1: Summary statistics of the three source files")
```

The three files together contain over 4 million lines and roughly 100 million words. Twitter entries are the shortest on average (~13 words per line), while blogs are the longest (~42 words per entry).

---

## 2. Sampling

Given the file sizes, a random sample of 1% per file is used for the exploratory analysis to keep computation manageable.

```{r sampling}
sample_pct <- 0.01

blogs_s   <- sample(blogs,   size = floor(length(blogs)   * sample_pct))
news_s    <- sample(news,    size = floor(length(news)    * sample_pct))
twitter_s <- sample(twitter, size = floor(length(twitter) * sample_pct))

all_text <- c(blogs_s, news_s, twitter_s)
```

---

## 3. Word Length Distribution

```{r word_length_hist, fig.width=8, fig.height=4}
word_lengths <- stri_length(unlist(stri_extract_all_words(all_text)))
word_lengths <- word_lengths[!is.na(word_lengths)]

ggplot(data.frame(len = word_lengths), aes(x = len)) +
  geom_histogram(binwidth = 1, fill = "#2c7bb6", color = "white", alpha = 0.85) +
  scale_x_continuous(limits = c(1, 20), breaks = 1:20) +
  labs(
    title = "Distribution of Word Lengths (combined sample)",
    x     = "Word length (characters)",
    y     = "Frequency"
  ) +
  theme_minimal(base_size = 13)
```

Most words are between 3 and 8 characters long, with a peak around 3-4 characters, consistent with the high frequency of short function words in English.

---

## 4. Line Length Distribution by Source

```{r line_length_hist, fig.width=8, fig.height=4}
df_lines <- data.frame(
  words  = c(
    stri_count_words(blogs_s),
    stri_count_words(news_s),
    stri_count_words(twitter_s)
  ),
  source = c(
    rep("Blogs",   length(blogs_s)),
    rep("News",    length(news_s)),
    rep("Twitter", length(twitter_s))
  )
)

ggplot(df_lines, aes(x = words, fill = source)) +
  geom_histogram(binwidth = 5, color = "white", alpha = 0.8, position = "identity") +
  facet_wrap(~source, scales = "free_y") +
  labs(
    title = "Words per Line by Source",
    x     = "Words per line",
    y     = "Count"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("Blogs" = "#d7191c", "News" = "#fdae61", "Twitter" = "#2c7bb6"))
```

Twitter entries cluster tightly below 30 words due to the character limit. News and blog entries show broader distributions, with some blog posts exceeding 200 words per paragraph.

---

## 5. Top Words (after removing stopwords)

```{r top_words, fig.width=8, fig.height=5}
sw <- tm::stopwords("en")

words_raw <- unlist(strsplit(tolower(paste(all_text, collapse = " ")), "[^a-z']+"))
words_raw <- words_raw[nchar(words_raw) > 1 & !words_raw %in% sw]

freq_table <- sort(table(words_raw), decreasing = TRUE)
top30 <- data.frame(
  word = names(freq_table[1:30]),
  freq = as.integer(freq_table[1:30])
)

ggplot(top30, aes(x = reorder(word, freq), y = freq)) +
  geom_bar(stat = "identity", fill = "#2c7bb6", alpha = 0.85) +
  coord_flip() +
  labs(
    title = "Top 30 Most Frequent Words (stopwords removed)",
    x     = "Word",
    y     = "Frequency"
  ) +
  theme_minimal(base_size = 12)
```

---

## 6. Plan for the Prediction Algorithm and Shiny App

**Algorithm:**  
The prediction model will use an **n-gram language model** (unigrams, bigrams, trigrams). The approach is:

1. Build frequency tables of n-grams from the full cleaned corpus.
2. Use a **backoff strategy** (Katz backoff or Stupid Backoff): if a trigram is not found, fall back to bigram, then unigram.
3. Return the top-3 most likely next words given the user's input.

**Shiny App:**  
A simple interface where the user types a partial sentence and the app suggests the next word in real time. The model will be pre-computed and stored as an `.rds` file for fast lookup.

**Key considerations:**

- Profanity filtering will be applied during corpus cleaning.
- Low-frequency n-grams (count < 3) will be pruned to reduce memory footprint.
- The app must respond in under 100ms, so efficiency in the lookup table is a priority.
